Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity

### 2022-01-21 manuscript_v7.Rmd
manuscript edits + related reruns

---
title: "Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via evasion. Despite much progress, there is still significant uncertainty in the temporal dynamics of gas exchange velocity (i.e. the speed of evasion) and thus carbon emissions to the atmosphere. We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring gas exchange velocity from future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'Chainsaw-eD',]$r2,2)*100`% of variation in gas exchange. We then couple this model with established Bayesian inference techniques to develop BIKER, or the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 763 discrete measurements of river gas exchange, yielding an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median normalized RMSE of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$NRMSE), 2)*100`%. Like many fluvial Bayesian inference algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), rather than estimating the absolute rates of exchange. Finally, we show BIKER robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER can be deployed to map the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and water turbulence), in addition to their long understood role of transporting carbon to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013] (*Liu etal in review- add number is paper if accepted in time*), with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.`r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])`%- @raymondGlobalCarbonDioxide2013], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

@@

To that end, the field has made significant progress in estimating $CO_2_{water}$, where we now have river-reach explicit modeling at monthly temporal resolutions using global hydrography datasets [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu etal in review*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

@@

To answer this first question, we use the specific hydraulic geometry, or the fundamental geomoprhic relationships between streamflow and channel shape [@leopoldHydraulicGeometryStream1953] of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange. We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any SWOT-observable river on Earth. Finally, we also test whether BIKER can predict actual gas fluxes when paired with an in situ $CO_2$ sensor and compare it against established methods that are reliant on in situ hydraulic measurements.

@@

## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in large rivers, and then obtain a suite of models we can empirically test. Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2). Following this model development, all models are empirically tested (section 2.3)

### 2.1 Data
Our gas exchange theory is developed using multiple datasets of directly measured $k_{600}$ and directly measured stream hydraulics collected from the literature. In total, we have collected 763 measurements of $k_{600}$, with 701 of these measurements previously published by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements were measured by @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements. All measurements come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale.

To qunatify how many SWOT-obsrvable rivers meet our hydraulic assumptions, we expand our set of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally taken to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.

### 2.2 Process-based modeling of river *k*
*k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. Based on classic surface-renewal theory, this model scales *k* via the smallest-scale (the 'Kolmogorov scale') turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. This model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Further, some workers have noted that laboratory and field observations suggest open channel flows with small bed roughness (generally big rivers) do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. *HERE: Zappa's note that this is frequently done in air-ocean gas exchange studies*. This is equation 4 and is referred to as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and lead to gas exchange [@talkeTurbulentKineticEnergy2013]. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum. Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a hydraulically-wide river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
Finally, both equations rely on measurements of $\epsilon$, of which measuring directly is non-trivial. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall such that $\epsilon = \epsilon_S = \frac{U_*^3}{H}$ [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993]. Another approach models *k* via 'form-drag dissipation' ($\epsilon = \epsilon_D = gS\bar{U}$) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that this model can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

### 2.3 Imposing large-river hydraulic geometry on *k* models
Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the distinct hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$. This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average $\frac{R_h}{H}$ ratio across our `r HG_swot$n` SWOT-observable river flows is `r HG_swot$mean * 100`% (see Text S1 for how we built this dataset). We refer to these flows as 'hydraulically-wide'.

Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models already outlined. While the complete model derivations for the small-eddy models (section 2.3.1) and the chainsaw models (section 2.3.2) are provided in Text S2, the broad goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.

#### 2.3.1 'Small-eddy' model for gas evasion
We derive equations starting from the small-eddy model (equation 3) for hydraulically-wide channels. If we assume $\epsilon=\epsilon_S$ and $R_H = H$ in equation 2, and finally solve for $k_{600}$ via statistical parameter $\alpha$, we eventually arrive at equation 5. If we do the same but assume $\epsilon=\epsilon_D$ and statistical parameter $\alpha_1$, we arrive at equation 6.

@@

#### 2.3.2 'Reynolds extension' model for gas evasion
We also impose $\epsilon=\epsilon_S$ and $R_H=H$ on the Reynolds extension model just like we did in section 2.3.1 and arrive at equation 7 (with statistical coefficient $\beta$). The result is identical to the hydraulically-wide small-eddy model (equation 5), except for the exponents which are both steeper. Equation 8 reflects the same derivation, but when $\epsilon=\epsilon_D$ and statistical coefficient $\beta_1$.

$$k_{600}=\beta(gS)^{9/16}H^{11/16} \mathbf{(7)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(8)}$$
We now have four different models for $k_{600}$ in hydraulically-wide rivers, each with a different set of assumptions and parameters but all defined as functions of the water surface slope and other hydraulic terms that are not directly measurable via SWOT. All four models are summarized in Table 1.

|**Description**|**Equation**|**Empirical test**|**Empirical fit via $r^2$**|
|---------------|------------|------------------|-----------------|
|Small-eddy, log-law-of-the-wall|$k_{600}=\alpha(gs)^{3/8}H^{1/8}$|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
|Small-eddy, form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
|Chainsaw, log-law-of-the-wall|$k_{600}=\beta(gs)^{9/16}H^{11/16}$|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)`|
|Chainsaw, form-drag dissipation|$k_{600}=\beta_1(gs)^{7/16}\bar{U}^{1/4}H^{9/16}$||Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)`|

### 2.4 Model validation
With equations 5-8 derived, we now empirically test their strength of fit for hydraulically-wide river flows. We validate on the dataset of direct measurements of $k_{600}$ and stream hydraulics, after filtering for measurements made in hydraulically-wide channels (n= `r n`). As equations 5-8 necessitate that this assumption is true, we defined 'hydraulically-wide' as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with. Models were assessed via the coefficient of determination ($r^2$) and results are plotted in Figure 3. Note that Figure 3 axes are plotted in logarithmic space just for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

![Figure 2: Empirical testing of four models for predicting gas exchange velocity in hydraulically-wide channels: (a) Small-eddy model with log-law-of-the-wall dissipation. (b) Chainsaw model with log-law-of-the-wall dissipation. (c) Small-eddy model with form-drag dissipation. (d) Chainsaw model with form-drag dissipation. Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg)

Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c). They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c). However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively). Modeling dissipation via total form-drag rather than just via bed shear and the log-law-of-the-wall improves model performance. This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].

The success of the Reynolds extensions model in hydraulically-wide channels, and in particular when coupled with the $\epsilon=\epsilon_D$ model (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 8 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

>>>>

---
title: "Using hydraulic geometry and the SWOT satellite to remotely sense river gas exchange velocity"
author:
  - Craig B Brinkerhoff:
      email: cbrinkerhoff@umass.edu
      institute: UMass
      correspondence: true
  - Colin J Gleason:
      institute: UMass
  - Christopher J Zappa:
      institute: Lamont
  - Peter A Raymond:
      institute: Yale
  - Merritt H Harlan:
      institute: UMass
institute:
  - UMass: Department of Civil & Environmental Engineering, University of Massachusetts-Amherst
  - Lamont: Lamont-Doherty Earth Observatory, Columbia University
  - Yale: School of the Environment, Yale University
bibliography: References.bib
csl: agu_citation_style.csl
output:
  redoc::redoc:
#output:
#  word_document:
      pandoc_args:
        - '--lua-filter=scholarly-metadata.lua'
        - '--lua-filter=author-info-blocks.lua'
      reference_docx: word-styles-reference.docx
---

@@

## Abstract
Extensive research over the past two decades has shown that the global river network emits significant amounts of greenhouse gas via evasion. Despite much progress, there is still large uncertainty in the temporal dynamics of gas exchange velocity and thus carbon emissions to the atmosphere. We propose that the NASA/CNES/USKA/CSA SWOT satellite, set to usher in a new age of hydrology research at the global scale, can provide similarly transformative insights to fluvial gas exchange modeling upon launch in 2022. Here, we present work inferring gas exchange velocity from future SWOT observations without in situ calibration. We exploit the unique geomorphology of SWOT-observable rivers to develop a physical model of gas exchange that is nearly entirely remotely sensible and explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of variation in gas exchange. We then couple this model with established Bayesian inference techniques to develop BIKER, or the 'Bayesian Inference of the $k_{600}$ Exchange Rate' algorithm. We validate BIKER on `r nrow(results_k600_rivs[results_k600_rivs$errFlag == 0,])` SWOT-simulated rivers and 166 discrete measurements of river gas exchange, yielding an algorithm that predicts this theoretical model solely from SWOT observations with a by-river median Kling-Gupta Efficiency of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$KGE), 2)`. Like many similar Bayesian remote sensing algorithms, BIKER is better at reproducing the temporal dynamics of gas exchange (median correlation coefficient of `r round(median(results_k600_rivs[results_k600_rivs$errFlag==0,]$r), 2)`), than the absolute rates of exchange. Finally, we show BIKER robust to measurement errors implicit in the SWOT data. With SWOT set to launch in late 2022, we suggest that BIKER will be useful in mapping the global-scale spatiotemporal dynamics of fluvial gas exchange in large rivers.

## 1 Introduction
Natural systems play a critical role in the budgeting and accounting of the global carbon cycle under climate change. Following @colePlumbingGlobalCarbon2007, the global river network is recognized to emit substantial amounts of carbon to the atmosphere via evasion (gas exchange driven by a diffusion gradient and water turbulence), in addition to their long understood role of transporting carbon to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-2000 Tg C/yr [*Liu accepted*; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite rivers' incredibly small percentage of the global land surface [0.67%- *Liu accepted*], this C flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011].

@@

To that end, the field has made significant progress in estimating $CO_2_{water}$, where we now have river-reach explicit modeling at monthly temporal resolutions using global hydrography datasets [@horgbyUnexpectedLargeEvasion2019; @brinkerhoffLakeMorphometryRiver2021; @saccardiImprovingPredictionsStream2021; *Liu accepetd*], but an equivalently sophisticated representation of $k_{CO_2}$ is still lacking. As equations 1 and 2 dictate, $CO_2_{water}$ and $k_{CO_2}$ share the burden of calculating $FCO_2$ and therefore these next-generation models will have to contend with this discrepancy, which is directly dictated by the lack of direct measurements of global river hydraulics.

@@

To answer this first question, we use the hydraulic geometry [the fundamental geomorphic relationships between streamflow and channel shape- @leopoldHydraulicGeometryStream1953] of large, SWOT-observable rivers to develop a process-based model for large-river gas exchange. We then take these findings and explore question 2 by implementing the model within an RS of $k_{600}$ algorithm named BIKER ('Bayesian Inference of the $k_{600}$ Evasion Rate'). The goal with BIKER is to require no in situ inputs of any kind (although in situ data could be ingested and should improve results) such that it is completely globally implementable on any SWOT-observable river on Earth. Finally, we also test whether BIKER can predict actual gas fluxes when paired with an in situ $CO_2$ sensor and compare it against established methods that are reliant on in situ hydraulic measurements.

@@

## 2 Hydraulic geometry in large rivers and predicting fluvial gas exchange velocity
To predict *k* in the very large rivers that SWOT will observe, we start from established process-based models for *k*, impose hydraulic assumptions that are valid in SWOT-observable rivers, and then obtain a suite of models we empirically test. Following a description of the data used (section 2.1), we outline these established models (section 2.2) and then step through our hydraulic assumptions to arrive at equations that can be ingested by SWOT (section 2.2). Following this model development, all models are empirically tested (section 2.3)

### 2.1 Data
We develop our gas exchange theory using multiple datasets of directly measured $k_{600}$ and directly measured stream hydraulics collected from the literature. In total, there are 763 measurements of $k_{600}$, with 701 of these measurements previously gathered by @raymondScalingGasTransfer2012 and @ulsethDistinctAirWater2019. The remaining measurements come from @churchillPredictionStreamReaeration1964 and @owensReaerationStudiesStreams1964. See Table S1 for a complete list of the studies that collected these measurements. All measurements come from tracer studies and thus define $k_{600}$ and the rivers' mean hydraulic properties at the reach scale.

To quantify how many SWOT-observable rivers meet our hydraulic assumptions, we expand our set of stream hydraulics (but not $k_{600}$) using a previously published compilation of in situ hydraulic measurements [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, and discharge from across the continental United States, originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and made public by the USGS.

### 2.2 Process-based modeling of river *k*
*k* should scale with near-surface turbulence in turbulent streamflows [@hallGasExchangeStreams2020] and thus extensive field and laboratory experiments have converged on the 'small-eddy model' as derived by @lamontEddyCellModel1970 and empirically anticipated by @calderbankContinuousPhaseHeat1961. This model scales *k* via the smallest-scale turbulent eddies and has been repeatedly empirically validated in freshwater systems [e.g. @katulStructureFunctionModel2018; @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @vachonRelationshipNearsurfaceTurbulence2010; @tokoroHighGastransferVelocity2008; @wangPhysicallyBasedScaling2021]. The small-eddy model is provided as equation 3, where $\epsilon$ is the dissipation rate of near-surface turbulence, $v$ is the kinematic viscosity, and *Sc* is the Schmidt number.

$$k\propto Sc^{-1/2}(v\epsilon)^{1/4} \mathbf{(3)}$$
Some laboratory and field observations additionally suggest that open channel flows with small bed roughness do not exhibit homogeneous surface dissipation at their air-water interface [@moogAirWaterGasTransfer1999; @talkeTurbulentKineticEnergy2013]. Given this observation, @moogAirWaterGasTransfer1999 proposed an extension to the small-eddy model, additionally scaling $k_{600}$ using a shear Reynold's number formulation. This is equation 4 and is referred to here as the 'Reynolds extension' model. The Reynolds model is hypothetically useful in low-turbulence flows where a relative lack of large-scale eddies effectively 'filter out' the number of small-eddies that actually reach the interface and initiate gas exchange [@talkeTurbulentKineticEnergy2013]. Scaling *k* via a shear Reynold's formulation is sometimes done to parameterize wave-breaking gas exchange models in the open ocean  [@brumerWaveRelatedReynoldsNumber2017; @zhaoDependenceWhitecapCoverage2001; @zhaoEffectWindWaves2003], though is infrequently done in rivers. In the context of BIKER, we chose to test this model because large, SWOT-observable rivers are generally the smoothest, least-turbulent flows along the stream-to-ocean continuum. Further, to our knowledge this Reynolds extension model has never been empirically tested in predicting river *k*, aside from confirming that large-scale eddies differentially move turbulence to the surface in a large river [@talkeTurbulentKineticEnergy2013].

$$k \propto Sc^{-1/2}(v\epsilon)^{1/4}(\frac{U_*H}{v})^{3/8} \mathbf{(4)}$$
Equations 3 and 4 both rely on $\epsilon$, of which measuring directly is non-trivial. When working at large scales, a commonly used $\epsilon$ model assumes that all turbulence is generated at the bed and transported to the air-water interface via the log-law-of-the-wall such that $\epsilon = \epsilon_S = \frac{U_*^3}{H}$ [@lorkeUnifiedScalingRelation2006; @nezuTurbulenceOpenChannel1993]. Another approach models *k* via 'form-drag dissipation' ($\epsilon = \epsilon_D = gS\bar{U}$) which is equivalently the total stream power per unit mass water. This normalized stream power captures the bulk frictional resistance (and thus energy dissipation) via channel banks, meanders, bars, etc. that is unique to fluvial systems [@moogStreamReaerationNonuniform1999]. Authors have since shown that this model can reasonably predict *k* in rivers and streams [@raymondScalingGasTransfer2012; @ulsethDistinctAirWater2019; @wangPhysicallyBasedScaling2021].

### 2.3 Imposing large-river hydraulic geometry on *k* models
Given the theoretical context privded in Section 2.2, we know turn to SWOT-observable systems specifically. Rivers and streams change predictably along their longitudinal profile from headwater to ocean, and we can exploit the hydraulic geometry of large rivers at the end of this continumn to estimate *k* in SWOT-observable systems. In general, as river size increases, channels become more rectangular, their shapes elongate (becoming wider quicker than they become deeper) and their hydraulic radii begin to approximate their mean flow depth, i.e. $R_h \approx H$. This is a common assumption in hydraulic and geomorphic modeling of large rivers, where the average SWOT-observable river flow has a $\frac{R_h}{H}$ ratio of `r HG_swot$mean * 100`% (n = `r HG_swot$n`; see Text S1 for how we built this dataset). We refer to these flows as 'hydraulically-wide'.

Assuming that all SWOT-observable rivers are hydraulically-wide, we test the performance of four models for predicting *k* in hydraulically-wide channels: the four unique combinations of equations 3-4 and the two $\epsilon$ models outlined above. While the complete model derivations for the small-eddy models (section 2.3.1) and the reynolds-extension models (section 2.3.2) are provided in Text S2, the final equations are provided in Table 1. The overall goal was to reduce the equations down to their fundamental parameters, identifying which terms are SWOT observable and limiting the number of terms not directly measurable via SWOT.

|**$k_{600}$ model**|**$\epsilon$ model**|**Final equation**|**Empirical test**|**Empirical fit ($r^2$)**|
|-----------|--------------------|------------------|-----------------|----------------------------|
|Small-eddy|Log-law-of-the-wall|$k_{600}=\alpha(gS)^{3/8}H^{1/8}$|Figure 3a|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)`|
|Small-eddy|Form-drag dissipation|$k_{600}=\alpha_1(gS\bar{U})^{1/4}$|Figure 3c|`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)`|
|Reynolds-extension|Log-law-of-the-wall|$k_{600}=\beta(gS)^{9/16}H^{11/16}$|Figure 3b|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)`|
|Reynolds-extension|Form-drag dissipation|$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16}$|Figure 3d|`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)`|

#### 2.3.1 'Small-eddy' model for gas evasion
We derive equations starting from the small-eddy model (equation 3) for hydraulically-wide channels. If we assume $\epsilon=\epsilon_S$ and $R_H = H$ in equation 2, and solve for $k_{600}$ via statistical parameter $\alpha$, we eventually arrive at equation 5. If we do the same but assume $\epsilon=\epsilon_D$ and statistical parameter $\alpha_1$, we arrive at equation 6.

@@

#### 2.3.2 'Reynolds extension' model for gas evasion
We also impose $\epsilon=\epsilon_S$ and $R_H=H$ on the Reynolds extension model just like we did in section 2.3.1 and arrive at equation 7 (with statistical coefficient $\beta$). The result is identical to the hydraulically-wide small-eddy model (equation 5), except for the exponents which are both steeper. Equation 8 reflects the same derivation, but with $\epsilon=\epsilon_D$ and statistical coefficient $\beta_1$.

$$k_{600}=\beta(gS)^{9/16}H^{11/16} \mathbf{(7)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(8)}$$
We now have four different models for $k_{600}$ in hydraulically-wide rivers, each with a different set of assumptions and parameters but all defined as functions of the water surface slope and other hydraulic terms that are not directly measurable via SWOT. Again, all four models are summarized in Table 1.

### 2.4 Model validation
With equations 5-8 derived, we now empirically test their strength of fit for hydraulically-wide river flows. We validate on the dataset of in situ measurements of $k_{600}$, after filtering for measurements made in hydraulically-wide channels (n= `r n`), which was operationally defined as flows whose hydraulic radius was within 1% of their mean flow depth. All told, this amounts to `r n` direct measurements of hydraulically-wide $k_{600}$ and stream hydraulics to test with. Models were assessed via the coefficient of determination ($r^2$) and results are plotted in Figure 3. Note that Figure 3 axes are plotted in logarithmic space only for visualization: model fit and validation (via $r^2$) were calculated in linear space as their models dictate.

![Figure 2: Empirical testing of four models for predicting gas exchange velocity in hydraulically-wide channels: (a) Small-eddy model with log-law-of-the-wall dissipation. (b) Reynolds-extension model with log-law-of-the-wall dissipation. (c) Small-eddy model with form-drag dissipation. (d) Reynolds-extension model with form-drag dissipation. Note that axes are plotted in logarithmic space just for visualization: model fit and validation were calculated in linear space.](cache/k600_theory/k600Plot.jpg){width=1000px}

Despite reasonable $r^2$ scores, both small-eddy models produce clearly biased estimates (Figures 3a, 3c). They systematically underestimate lower gas exchange velocities and overestimate higher ones, with the log-law-of-the-wall model (Figure 3a) performing marginally worse both visually and via $r^2$ (`r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eS',]$r2, 2)` versus `r round(modelsWIDE[modelsWIDE$name == 'Small-eddy-eD',]$r2, 2)` for Figure 3c). However, when this model is expanded to retard the dissipation rate in the smoothest flows via the Reynolds extension model [equations 6 and 7- @moogAirWaterGasTransfer1999], the dynamics are more accurately captured (Figures 3b, 3d) and $r^2$ scores go up appreciably (`r round(modelsWIDE[modelsWIDE$name == 'reynolds-eS',]$r2, 2)` and `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2, 2)` for the log-law-of-the-wall and form-drag dissipation, respectively). Modeling dissipation via total form-drag rather than just via bed shear improves model performance regardless of $k_{600}$ equation used. This is likely because the majority of dissipation occurring in rivers comes not just from the streambed, but from bulk form drag via channel walls, bars, and meanders [@raymondScalingGasTransfer2012; @moogStreamReaerationNonuniform1999; @wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019].

The success of the Reynolds-extension models in hydraulically-wide channels, and in particular when coupled with form-drag dissipation (equation 8, Figure 3d), provides us with a strong physical-model for gas evasion. The river hydraulics terms in equation 8 can either be directly measured or reasonably inferred from SWOT measurements, effectively opening the door for remotely sensing the gas exchange velocity. This is explored next.

## 3 Exploiting equation 8 to remotely sense gas exchange velocity
We have shown that scaling $k_{600}$ via equation 8 explains `r round(modelsWIDE[modelsWIDE$name == 'reynolds-eD',]$r2,2)*100`% of the variation in $k_{600}$ in hydraulically-wide rivers, which includes nearly all SWOT-observable rivers. Further, equation 8 has only three non-remotely-sensible terms: $k_{600}$, mean flow depth, and mean flow velocity. Conveniently, techniques to simultaneously infer hydraulic parameters from SWOT data have been established over the last decade to infer streamflow from SWOT's measurements [e.g. @andreadisConstrainingAssimilationSWOT2020; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. For BIKER, we follow the work developed by @hagemannBAMBayesianAMHGManning2017, @brinkerhoffConstrainingRemoteRiver2020, and @durandEstimatingReachaveragedDischarge2014 to infer $k_{600}$, channel bathymetry, and roughness from SWOT observations using a modified form of Manning's equation for mean flow velocity. Following a description of the data used (section 3.1), we detail algorithm development and experimental design (section 3.2) and then we present the validation results (section 3.3).
